Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-07 paper.tex
algo details

% history of approximating the coalescent: 
SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
(2000)
These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
Clear distinction into two groups: ARGWeaver infers actual ARG.
Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
structures. (other peculiarities: ...).

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
\caption{Description of algorithm}
\end{figure}

@@

The likelihood computation as described above can be done with a single pass 
of the edge table.
The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
with at time $t$. 
The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
that we can impose a strict total order on all (overlapping) lineages based on their 
starting point.

The tree sequence encoding provides us with a natural order on all lineages.  
For each local tree with left coordinate $x$, lineages associated with an edge 
starting at $x$ can only coalesce with the subset of lineages associated with an 
edge that is represented in that same local tree. Because local tree changes 
only occur whenever edges are removed and/or inserted,
we can rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

% edge diff algorithm
The order in which edges are inserted and removed is determined by the index 
vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}. The edge insertion vector 
$\textbf{i}$ gives the ordering of edges sorted by left endpoint, and among edges 
with the same left endpoint, sorted so that edges closer to the root appear later. 
The edge removal vector $\textbf{o}$ is similar, except gives the ordering of edges 
by right endpoint and with edges closer to the root appearing sooner. This order imposes the 
strong ordering on all edges as required by the definition of $I_{kl}$. The variables $\texttt{i}$ and 
$\texttt{o}$ variables are used to keep track of our position in the edge insertion and edge removal 
indexes.\\

% the counting algorithm
To transition from one tree to the next, we first iterate across the indices of all edges 
in $\textbf{o}$ from its current index up to $\texttt{o}$. For each edge in we decrement $I$ 
from index \texttt{edge.child} to \texttt{edge.parent}. We subsequently iterate across the 
edges in $\textbf{i}$ up to index $\texttt{i}$. For each new edge, we first compute the likelihood 
associated with that edge and then 
increment $I$ from \texttt{edge.child} to \texttt{edge.parent}.\\

% improve this section, then transition to limitations of unary nodes
We further rely on the assumption that the unary (coalescent) nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm
However,

@@

\subsection{Flexibility}

By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence.

The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
and simplification.
Similarly, any subset of the tree sequence that is itself a valid tree sequence can again 
be used to compute a likelihood. Such operations include: slicing genome both in time, and space.

Strength of ARGs is the ability to gain temporal resolution. This approach allows us to fully 
benefit from this.

We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).

@@

%\subsubsection{Node labellings}
% implement idea from Guo et al. 2022.

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. 
Important step moving forwards towards bridging the gap between the tremendous 
recent advances in ARG inference and current state of the art statistical inference of 
past demography and selection (but see). % some/many of these approaches 
% ignore recombination

>>>>

% history of approximating the coalescent: 
SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly 
(2000)
These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
Clear distinction into two groups: ARGWeaver infers actual ARG.
% what does kwarg do?
Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
structures. (other peculiarities: ...).

% computing likelihoods on ARGs; from Kuhner and Felsenstein to ARGWeaver

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
\caption{Description of algorithm: the algorithm progresses from left to right and 
keeps track of the . Let's assume we have visited the first 5 edges present in the 
first tree (tabel 1). Moving onto the next tree, need to remove single edge 
$(2, 5, [0,42))$, update the counts $I$ by decrementing all intervals between 5 and 
2. The state of the vector is now $I_6$, the number of extant lineages the lineage 
associated with edge $(1, 6, [42, 100)$ can coalesce with. Now compute the likelihood 
associated with this edge. Then insert this edge and increment the count vector between 
1 and 6. Two intermediate steps to move to the second tree. Similarly again two steps 
to move to the last tree and insert edge $(5, 7, [70,100))$
}

\end{figure}

@@

The likelihood computation as described above can be done with a single pass 
of the edge table.
The central problem however is to efficiently maintain the state $I_e(t)$, which counts 
the number of lineages (the lineage associated with) the focal edge $\epsilon$ can coalesce 
with at time $t$. 
The definition of $I_{kl}$ (\ref{def:coal}) is based on the assumption 
that we can impose a strict total order on all (overlapping) lineages based on their 
starting point. The tree sequence encoding provides us with a natural order on all lineages.  
For each local tree with left coordinate $x$, lineages associated with an edge 
starting at $x$ can only coalesce with the subset of lineages associated with one of the
edges that make up that local tree. Each of these edges will either already have 
been present in the previous local tree, or will have the same starting point $x$.
In \texttt{tskit}, the order in which edges are inserted and removed 
while moving from one tree to the next, is determined by the index 
vectors $\textbf{i}$ and $\textbf{o}$ \citep{kelleher_efficient_2016}.
The edge insertion vector $\textbf{i}$ gives the ordering of edges sorted by left 
endpoint, and among edges with the same left endpoint, sorted so that edges closer 
to the root appear later. The edge removal vector $\textbf{o}$ is similar, 
except gives the ordering of edges by right endpoint and with edges closer to the 
root appearing sooner. The variables $\texttt{i}$ and 
$\texttt{o}$ are used to keep track of our position in the edge insertion and edge removal 
indexes as we move along the tree sequence. This order imposes the 
strong ordering on all edges as required by the definition of $I_{kl}$.\\

% edge diff algorithm
The size of the subset of lineages in each internode interval each lineage 
can coalesce with can thus be maintained by keeping track of a single vector $I$ 
that is decremented or incremented as defined by
$\textbf{o}[\texttt{o}]$ and $\textbf{i}[\texttt{i}]$, 
Further keep track of a last-parent array to detect recombination events.
We can thus compute a likelihood for the tree sequence, while visiting each edge only once.

More concretely, prior to inserting an edge, $I$ represents the number of lineages 
the lineage associated with 
that edge can coalesce with. Then insert the edge and update the counts.

To transition from one tree to the next, we first iterate across the indices of all edges 
in $\textbf{o}$ from its current index up to $\texttt{o}$. For each edge in we decrement $I$ 
from index \texttt{edge.child} to \texttt{edge.parent}. We subsequently iterate across the 
edges in $\textbf{i}$ up to index $\texttt{i}$. For each new edge, we first compute the likelihood 
associated with that edge and then 
increment $I$ from \texttt{edge.child} to \texttt{edge.parent}.\\

% improve this section, transition to modifications in the algorithm in case of 
% incomplete unary nodes, polytomies, ...
We further rely on the assumption that the unary (coalescent) nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm

@@

\subsection{Flexibility / general class of ARGs}

% highlight in what ways we can now compute likelihoods for a GENERAL class of ARGs

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with.

Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample.

% something on the process backwards in time
We can readily incorporate changes in $N_e$ and $r$, in both dimensions (time and along the genome).

@@

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. 
Important step moving forwards towards bridging the gap between the tremendous 
recent advances in ARG inference and current state of the art statistical inference of 
past demography and selection (but see ). % some/many of these approaches

% ignore recombination
